---
title: "THE EFFECT OF HARMONICITY ON MISMATCH NEGATIVITY
RESPONSES TO DIFFERENT AUDITORY FEATURES - power analysis"
author: "Krzysztof Basiński (based on a script by David Quiroga)"
date: '2024-12-09'
output: 
  pdf_document: 
    keep_tex: yes
---

First, load the _lme4_ library and define a function to conduct power estimates.

```{r}

library(lme4)

pwr_calc_test <- function(nsims, ssizes) {
  # intialize data frame to store the output 
  pwr <- data.frame() 
  
  for (s in 1:length(ssizes)) {
    for (n in 1:nsims) {
      # df index
      idx <- (s-1) * nsims + n
      
      # current sample size
      ssize <- ssizes[s]
      
      # conditions factor
      conds <- rep(c(0,1,2), each = 5, times =1)
      
      #feature factor
      feature <- rep(c(1:5), times = 3)

      make_ziutek <- function(sub_id, params) {
        
        # add intercept for condition
        intercept_cond <- rep(rnorm(3,params$b0c, params$b0c_sd), each = 5)

        # add feature intercept
        intercept_feature <- rep(rnorm(5, params$b0f, params$b0f_sd), times=3)
        
        # add condition and feature effects
        beta_cond <- params$b1_c * conds
        beta_feature <- params$b1_f * feature
        
        # add interaction
        beta_interaction <- params$b1_i * conds * feature
        
        # add residual noise
        residuals <- rnorm(length(conds), 0, params$residuals_sd)

        # collect in a dataframe and calculate simulated measured outcome (y)
        d <- data.frame('cond' = as.character(conds), 
                        'feature' = as.character(feature),
                        'sub' = sub_id,
                        'b0_c' = intercept_cond,
                        'b0_f' = intercept_feature,
                        'b1_c' = beta_cond,
                        'b1_f' = beta_feature,
                        'res' = residuals,
                        'y' = intercept_cond + intercept_feature + beta_cond + beta_feature + beta_interaction + residuals)
        }
      
      ziutek_list <- lapply(1:ssize, make_ziutek, params)
      d <- do.call(rbind, ziutek_list)
      
      # fit models - make sure they're correct
      m0 <- lmer(y ~ cond+feature + (1|sub), data = d, REML = FALSE)
      m1 <- lmer(y ~ cond*feature + (1|sub), data = d, REML = FALSE)
      
      # perform likelihood ratio test
      test <- anova(m0,m1)
     
       #store output of simulation
      pwr[idx,'sim'] <- n
      pwr[idx, 'ssize'] <- ssize
      pwr[idx, 'b0'] <- summary(m1)$coefficients[1]
      pwr[idx, 'b1'] <- summary(m1)$coefficients[2]
      pwr[idx, 'sd_int'] <- attr(summary(m1)$varcor$sub,"stddev") 
      pwr[idx, 'sd_res'] <- summary(m1)$sigma
      pwr[idx, 'x2'] <- test$Chisq[2]
      pwr[idx, 'p'] <- test$`Pr(>Chisq)`[2]
    }
  }
  return(pwr)
} 

```

This function uses a mixed effects modelling to compare a null model in the form:

` y ~ harmonicity + (1 | participant) `

to the model:

` y ~ harmonicity*feature + (1 | participant) `

Where `y` is the dependent variable (EEG results), harmonicity is a fixed effect and participant is a random effect. Harmonicity is specified as a three-level factor (harmonic, inharmonic, inharmonic changing).


Now specify the parameters to simulate EEG data. Let's assume a minimum difference between conditions of -1uV (micro volts) and a residual SD of 1.5uV. This script runs 2000 simulations for each possible sample size from N = 10 to N = 30.

```{r message=FALSE, warning=FALSE}

# parameters
params <- list(
    b0c = 1,
    b0c_sd = 1,

    b0f = 6,
    b0f_sd = 1,

    b1_c = 1,
    b1_f = 3,
    b1_i = .3,
    residuals_sd = 1
)

nsims <- 1000 # number of simulations per sample size
ssizes <- c(5, 10, 20, 30) # sample sizes

set.seed(777)
pwr_test <- pwr_calc_test(nsims, ssizes)

summary2 <- aggregate(pwr_test$p, by = list(pwr_test$ssize), FUN = function(x) sum(x < 0.05)/length(x))
colnames(summary2) <- c('sample.size','power')
print(summary2)


```


Now plot the power curve as a function of sample size:
```{r warning=FALSE}
with(summary2, plot(sample.size, power, type = 'ol')) 
abline(h=.8, col='red')
abline(h=.9, col='green')
title('Power curve')
```

This analysis shows that a power level of 0.8 is achieved for a sample size of N = 12 and power level of 0.9 for a sample size of N = 16.

Now repeat the process for a weaker effect of -.8uV:

```{r message=FALSE, warning=FALSE}

e3 <- -.8 # -1 uv # uV # minimum difference between conditions

set.seed(777)
pwr2 <- pwr_calc(e0,e3,e0_sd,eres_sd,nsims,ssizes)

summary3 <- aggregate(pwr2$p,by = list(pwr2$ssize), FUN = function(x) sum(x < 0.05)/length(x))
colnames(summary3) <- c('sample.size','power')
print(summary3)

```

And plot the power curve:

```{r warning=FALSE}
with(summary3, plot(sample.size, power, type = 'ol')) 
abline(h=.8, col='red')
abline(h=.9, col='green')
title('Power curve')
```

Here, satisfactory statistical power of 0.8 is achieved above N = 18, while N = 23 provides power above 0.9.

This analysis indicates that the planned sample size of N = 30 will provide enough statistical power to reliably detect a difference between conditions of 0.8uV.